---
title: "matrix_analysis"
author: "Kexin Gong"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

0. Load required libraries and set a random seed
```{r}
library(tidyverse)
library(ggplot2)
library(uwot)
library(cluster)
library(pheatmap)
library(randomForest)
library(caret)
library(ggpubr)

set.seed(929) # Ensure reproducibility of the analysis results
```

1. Import data
```{r}
expr_df <- read.csv("/Users/kexin/Desktop/project/data/SCs_32_matrix.csv", check.names = FALSE)

dim(expr_df)
head(expr_df[, 1:6])

cell_labels <- expr_df$cell_label
```

2. Prepare expression matrix
```{r}
expr_mat <- expr_df %>% select(-cell_label)

# remove Blank probes
expr_mat <- expr_mat %>% select(!starts_with("Blank"))

# remove genes with NA
expr_mat <- expr_mat[, colSums(is.na(expr_mat)) == 0]

# remove zero-variance genes
gene_var <- apply(expr_mat, 2, var)
expr_mat_clean <- expr_mat[, gene_var > 0]

dim(expr_mat_clean)
```

3. Basic QC
```{r, fig.width=6, fig.height=4.5}
# Calculate basic quality control metrics for each cell
qc_df <- data.frame(
  cell_label = cell_labels,
  detected_genes = rowSums(expr_mat_clean > 0), # Number of genes detected per cell
  total_expression = rowSums(expr_mat_clean) # Total log2 expression per cell
)

# Plot the distribution of detected genes per cell
ggplot(qc_df, aes(x = detected_genes)) +
  geom_histogram(bins = 30) +
  theme_classic(base_size = 14) +
  labs(
    title = "Detected genes per cell",
    x = "Number of detected genes",
    y = "Cell count"
  )

# ggsave(filename = "/Users/kexin/Desktop/BINF-6999-Project/figures/SCs/detected_genes_per_cell.png", width = 6, height = 5, dpi = 300)

# Plot the distribution of total expression per cell
ggplot(qc_df, aes(x = total_expression)) +
  geom_histogram(bins = 30) +
  theme_classic(base_size = 14) +
  labs(
    title = "Total log2 expression per cell",
    x = "Total expression",
    y = "Cell count"
  )

# ggsave(filename = "/Users/kexin/Desktop/BINF-6999-Project/figures/SCs/total_expression_per_cell.png", width = 6, height = 5, dpi = 300)

```

4. PCA
```{r, fig.width=6, fig.height=5}
# Perform PCA on the cleaned expression matrix with centering and scaling
pca_res <- prcomp(expr_mat_clean, center = TRUE, scale. = TRUE)

# Convert PCA scores to a data frame and retain cell labels
pca_df <- as.data.frame(pca_res$x)
pca_df$cell_label <- cell_labels

# Calculate the percentage of variance explained by each principal component 
# and the cumulative variance explained
var_explained <- summary(pca_res)$importance[2, ] * 100
cum_var <- cumsum(var_explained)

# Generate axis labels showing the variance explained by PC1 and PC2
pc1_lab <- paste0("PC1 (", round(var_explained[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(var_explained[2], 1), "%)")

# Visualize cell distribution along the first two principal components
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs cells, section 32",
    x = pc1_lab,
    y = pc2_lab
  )
# ggsave(filename = "/Users/kexin/Desktop/BINF-6999-Project/figures/SCs/pca_scs_32.png", width = 6, height = 5, dpi = 300)

# Create a data frame containing PCA variance statistics
scree_df <- data.frame(
  PC_number = seq_along(var_explained),
  Variance = var_explained,
  Cumulative = cum_var
)

# Visualize the variance explained by the first 30 principal components
ggplot(scree_df[1:30, ], aes(x = PC_number, y = Variance)) +
  geom_point(size = 2) +
  geom_line() +
  theme_classic(base_size = 14) +
  labs(
    title = "Scree plot of PCA",
    x = "Principal component",
    y = "Variance explained (%)"
  )
# ggsave(filename = "/Users/kexin/Desktop/BINF-6999-Project/figures/SCs/pca_scree_32.png", width = 6, height = 5, dpi = 300)

# Visualize the cumulative variance explained by the first 30 principal components
ggplot(scree_df[1:30, ], aes(x = PC_number, y = Cumulative)) +
  geom_point(size = 2) +
  geom_line() +
  theme_classic(base_size = 14) +
  labs(
    title = "Cumulative variance explained by PCs",
    x = "Principal component",
    y = "Cumulative variance explained (%)"
  )
# ggsave(filename = "/Users/kexin/Desktop/BINF-6999-Project/figures/SCs/pca_cumulative_variance.png", width = 6, height = 5, dpi = 300)
```

5. Explore additional PCA dimensions
```{r, fig.width=6, fig.height=5}
ggplot(pca_df, aes(x = PC1, y = PC3)) +
  geom_point(size = 2.5, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs cells, section 32",
    subtitle = "PC1 vs PC3",
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC3 (", round(var_explained[3], 1), "%)")
  )

ggplot(pca_df, aes(x = PC2, y = PC3)) +
  geom_point(size = 2.5, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs cells, section 32",
    subtitle = "PC2 vs PC3",
    x = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    y = paste0("PC3 (", round(var_explained[3], 1), "%)")
  )

ggplot(pca_df, aes(x = PC1, y = PC4)) +
  geom_point(size = 2.5, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs cells, section 32",
    subtitle = "PC1 vs PC4",
    x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
    y = paste0("PC4 (", round(var_explained[4], 1), "%)")
  )

ggplot(pca_df, aes(x = PC2, y = PC4)) +
  geom_point(size = 2.5, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs cells, section 32",
    subtitle = "PC2 vs PC4",
    x = paste0("PC2 (", round(var_explained[2], 1), "%)"),
    y = paste0("PC4 (", round(var_explained[4], 1), "%)")
  )
```
6. Principal component loading analysis
```{r}
# Extract PCA loading matrix.
# Each loading represents the contribution of a gene to a principal component.

loadings <- as.data.frame(pca_res$rotation)
head(loadings)
```

6.1 PC1
```{r}
# Top positive genes contributing to PC1

pc1_positive <- loadings %>%
  rownames_to_column("gene") %>%
  arrange(desc(PC1)) %>%
  slice(1:20)

pc1_positive

# Top positive genes contributing to PC1

pc1_positive <- loadings %>%
  rownames_to_column("gene") %>%
  arrange(desc(PC1)) %>%
  slice(1:20)

pc1_positive

pc1_plot <- loadings %>%
  rownames_to_column("gene") %>%
  mutate(abs_loading = abs(PC1)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:20)

ggplot(pc1_plot,
       aes(x = reorder(gene, abs_loading),
           y = PC1)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    title = "Top gene loadings for PC1",
    x = "Gene",
    y = "Loading")

```

6.2 PC2
```{r}
# Top positive genes contributing to PC2

pc2_positive <- loadings %>%
  rownames_to_column("gene") %>%
  arrange(desc(PC2)) %>%
  slice(1:20)

pc2_positive

# Top negative genes contributing to PC2

pc2_negative <- loadings %>%
  rownames_to_column("gene") %>%
  arrange(PC2) %>%
  slice(1:20)

pc2_negative

pc2_plot <- loadings %>%
  rownames_to_column("gene") %>%
  mutate(abs_loading = abs(PC2)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:20)

ggplot(pc2_plot,
       aes(x = reorder(gene, abs_loading),
           y = PC2)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    title = "Top gene loadings for PC2",
    x = "Gene",
    y = "Loading")

```

6.3 PC3
```{r}
# Top positive genes contributing to PC3

pc3_positive <- loadings %>% 
  rownames_to_column("gene") %>%
  arrange(desc(PC3)) %>%
  slice(1:20)

pc3_positive

# Top negative genes contributing to PC3

pc3_negative <- loadings %>%
  rownames_to_column("gene") %>%
  arrange(PC3) %>%
  slice(1:20)

pc3_negative

pc3_plot <- loadings %>%
  rownames_to_column("gene") %>%
  mutate(abs_loading = abs(PC3)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:20)

ggplot(pc3_plot,
       aes(x = reorder(gene, abs_loading),
           y = PC3)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    title = "Top gene loadings for PC3",
    x = "Gene",
    y = "Loading")
```
7. Import metadata and add anatomical labels
```{r}
metadata <- read.csv(
  "/Users/kexin/Desktop/project/data/SCs_32.csv",
  check.names = FALSE
)

dim(metadata)
head(metadata[, 1:6])
colnames(metadata)

# Since the metadata and expression matrix are in the same cell order,
# we only check whether they contain the same number of cells.
stopifnot(nrow(metadata) == nrow(expr_df))

# Add metadata to PCA dataframe
pca_df$sample_id <- metadata$`Sample Id`
pca_df$cell_id <- metadata$`Cell Label`
pca_df$neurotransmitter <- metadata$Neurotransmitter.Type
pca_df$cell_class <- metadata$Class
pca_df$subclass <- metadata$Subclass
pca_df$supertype <- metadata$Supertype
pca_df$allen_cluster <- metadata$Cluster
pca_df$anatomical_structure <- metadata$Anatomical.Structure
pca_df$anatomical_substructure <- metadata$Anatomical.Substructure

# Check labels
table(pca_df$neurotransmitter)
table(pca_df$anatomical_substructure)
```

8. PCA colored by anatomical and biological labels
```{r, fig.width=6, fig.height=5}
# PCA colored by neurotransmitter type
ggplot(pca_df, aes(x = PC1, y = PC2, color = neurotransmitter)) +
  geom_point(size = 2.5, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs neurons colored by neurotransmitter type",
    x = pc1_lab,
    y = pc2_lab,
    color = "Neurotransmitter"
  )

# PCA colored by anatomical substructure
ggplot(pca_df, aes(x = PC1, y = PC2, color = anatomical_substructure)) +
  geom_point(size = 2.5, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs neurons colored by anatomical substructure",
    x = pc1_lab,
    y = pc2_lab,
    color = "Substructure"
  )

# PCA colored by subclass
ggplot(pca_df, aes(x = PC1, y = PC2, color = subclass)) +
  geom_point(size = 2.2, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "PCA of SCs neurons colored by Allen subclass",
    x = pc1_lab,
    y = pc2_lab,
    color = "Subclass"
  )
```

9. UMAP without clustering
```{r, fig.width=6, fig.height=5}
# Use the first 30 PCs for UMAP.
# No clustering is applied here. UMAP is used only to visualize transcriptomic structure.

pc_use <- 1:30
pca_input <- pca_res$x[, pc_use]

umap_res <- uwot::umap(
  pca_input,
  n_neighbors = 15,
  min_dist = 0.3,
  metric = "euclidean"
)

umap_df <- as.data.frame(umap_res)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$cell_label <- cell_labels
umap_df$sample_id <- metadata$`Sample Id`
umap_df$cell_id <- metadata$`Cell Label`
umap_df$neurotransmitter <- metadata$Neurotransmitter.Type
umap_df$cell_class <- metadata$Class
umap_df$subclass <- metadata$Subclass
umap_df$supertype <- metadata$Supertype
umap_df$allen_cluster <- metadata$Cluster
umap_df$anatomical_structure <- metadata$Anatomical.Structure
umap_df$anatomical_substructure <- metadata$Anatomical.Substructure

# UMAP without labels
ggplot(umap_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 2.5, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "UMAP of SCs neurons, section 32",
    subtitle = "UMAP based on PC1-PC30; no clustering applied",
    x = "UMAP1",
    y = "UMAP2"
  )
```

10. UMAP colored by anatomical and biological labels
```{r}
# UMAP colored by neurotransmitter type
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = neurotransmitter)) +
  geom_point(size = 2.5, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    title = "UMAP of SCs neurons colored by neurotransmitter type",
    subtitle = "UMAP based on PC1-PC30",
    x = "UMAP1",
    y = "UMAP2",
    color = "Neurotransmitter"
  )

# UMAP colored by anatomical substructure
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = anatomical_substructure)) +
  geom_point(size = 2.5, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    title = "UMAP of SCs neurons colored by anatomical substructure",
    subtitle = "Testing whether transcriptomic structure recovers SCs subregions",
    x = "UMAP1",
    y = "UMAP2",
    color = "Substructure"
  )

# UMAP colored by Allen subclass
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = subclass)) +
  geom_point(size = 2.2, alpha = 0.75) +
  theme_classic(base_size = 14) +
  labs(
    title = "UMAP of SCs neurons colored by Allen subclass",
    subtitle = "UMAP based on PC1-PC30",
    x = "UMAP1",
    y = "UMAP2",
    color = "Subclass"
  )
```

11. Distribution of anatomical substructures across Allen subclasses
```{r}
# Count the number of cells for each anatomical substructure
# within each Allen subclass.

subclass_table <- pca_df %>%
  count(anatomical_substructure, subclass)

subclass_table

# Calculate the percentage of each subclass within each anatomical substructure

subclass_percent <- subclass_table %>%
  group_by(anatomical_substructure) %>%
  mutate(percent = n / sum(n) * 100)

subclass_percent

ggplot(
  subclass_percent,
  aes(
    x = anatomical_substructure,
    y = percent,
    fill = subclass
  )
) +
  geom_col() +
  theme_classic(base_size = 14) +
  labs(
    title = "Allen subclass composition of SCs anatomical substructures",
    x = "Anatomical substructure",
    y = "Percentage of cells",
    fill = "Allen subclass"
  )
```

12. Neurotransmitter composition of SCs substructures
```{r}
neuro_table <- pca_df %>%
  count(
    anatomical_substructure,
    neurotransmitter
  ) %>%
  group_by(anatomical_substructure) %>%
  mutate(percent = n / sum(n) * 100)

neuro_table

ggplot(
  neuro_table,
  aes(
    x = anatomical_substructure,
    y = percent,
    fill = neurotransmitter
  )
) +
  geom_col() +
  theme_classic(base_size = 14) +
  labs(
    title = "Neurotransmitter composition across SCs substructures",
    x = "Anatomical substructure",
    y = "Percentage of cells",
    fill = "Neurotransmitter"
  )
```

13. Association between Allen subclasses and anatomical substructures
```{r}
subclass_table <- table(
  pca_df$anatomical_substructure,
  pca_df$subclass
)

subclass_table

chisq.test(
    subclass_table,
    simulate.p.value = TRUE,
    B = 10000
)
```

13.1 Standardized residual analysis
```{r, fig.width=7, fig.height=5}
chi_subclass <- chisq.test(
    subclass_table,
    simulate.p.value = TRUE
)

stdres <- chi_subclass$stdres

round(stdres,2)

pheatmap(
    stdres,
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    display_numbers = TRUE
)
```

14. Statistical comparison of PCA coordinates across anatomical substructures
```{r}
# Test whether PCA coordinates differ among SCs anatomical substructures.
# This evaluates whether transcriptomic variation is associated with anatomy.

pcs_to_test <- c("PC1", "PC2", "PC3", "PC4")

# Kruskal-Wallis tests for PC1-PC4
kruskal_results <- data.frame()

for (pc in pcs_to_test) {
  
  test_formula <- as.formula(
    paste(pc, "~ anatomical_substructure")
  )
  
  test_res <- kruskal.test(
    test_formula,
    data = pca_df
  )
  
  kruskal_results <- rbind(
    kruskal_results,
    data.frame(
      PC = pc,
      statistic = test_res$statistic,
      df = test_res$parameter,
      p_value = test_res$p.value
    )
  )
}

kruskal_results

# Pairwise Wilcoxon tests for each PC
pairwise_results <- list()

for (pc in pcs_to_test) {
  
  pairwise_results[[pc]] <- pairwise.wilcox.test(
    pca_df[[pc]],
    pca_df$anatomical_substructure,
    p.adjust.method = "BH"
  )
}

pairwise_results

comparisons <- list(
  c("SCop", "SCsg"),
  c("SCop", "SCzo"),
  c("SCsg", "SCzo")
)

# Visualize PC1 distribution across anatomical substructures.
# PC1 mainly separates SCop from the other two substructures.

ggplot(
  pca_df,
  aes(x = anatomical_substructure,
      y = PC1,
      fill = anatomical_substructure)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  stat_compare_means(
    comparisons = list(
      c("SCop","SCsg"),
      c("SCop","SCzo")
    ),
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme_classic(base_size = 14) +
  labs(
    title = "PC1 distribution across SCs anatomical substructures",
    x = "Anatomical substructure",
    y = "PC1 score"
  ) +
  guides(fill = "none")

# Visualize PC2 distribution across anatomical substructures.
# PC2 captures additional variation among the three substructures.

ggplot(
  pca_df,
  aes(x = anatomical_substructure,
      y = PC2,
      fill = anatomical_substructure)
) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.2, size = 1, alpha = 0.4) +
  stat_compare_means(
    comparisons = comparisons,
    method = "wilcox.test",
    label = "p.signif"
  ) +
  theme_classic(base_size = 14) +
  labs(
    title = "PC2 distribution across SCs anatomical substructures",
    x = "Anatomical substructure",
    y = "PC2 score"
  ) +
  guides(fill = "none")
```

15. Random forest classification of anatomical substructure
```{r}
# Prepare input expression matrix and labels.
# Rows are cells and columns are genes.
rf_x <- as.data.frame(expr_mat_clean)
rf_y <- as.factor(pca_df$anatomical_substructure)

# Split data into training and testing sets.
train_index <- createDataPartition(rf_y, p = 0.7, list = FALSE)

x_train <- rf_x[train_index, ]
x_test <- rf_x[-train_index, ]

y_train <- rf_y[train_index]
y_test <- rf_y[-train_index]

# Train random forest classifier.
rf_model <- randomForest(
  x = x_train,
  y = y_train,
  ntree = 500,
  importance = TRUE
)

rf_model

# Predict anatomical substructure for test cells.
rf_pred <- predict(rf_model, x_test)

# Confusion matrix shows how well expression predicts anatomy.
confusionMatrix(rf_pred, y_test)

# Extract top genes important for predicting anatomical substructure.

importance_df <- as.data.frame(importance(rf_model))
importance_df$gene <- rownames(importance_df)

top_rf_genes <- importance_df %>%
  arrange(desc(MeanDecreaseGini)) %>%
  slice(1:20)

top_rf_genes

ggplot(top_rf_genes, aes(x = reorder(gene, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    title = "Top genes predicting SCs anatomical substructure",
    x = "Gene",
    y = "Mean decrease in Gini"
  )
```

16. Summary of SCs anatomical substructure analysis
```{r}
# Summarize cell number, neurotransmitter composition,
# and mean PCA coordinates for each anatomical substructure.

scs_summary <- pca_df %>%
  group_by(anatomical_substructure) %>%
  summarise(
    n_cells = n(),
    mean_PC1 = mean(PC1),
    mean_PC2 = mean(PC2),
    mean_PC3 = mean(PC3),
    percent_GABA = mean(neurotransmitter == "GABA") * 100,
    percent_Glut = mean(neurotransmitter == "Glut") * 100
  )

scs_summary
```


