---
title: "matrix_analysis_sc"
author: "Kexin Gong"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

0. Load required libraries and set a random seed
```{r}
library(tidyverse)
library(ggplot2)
library(uwot)
library(cluster)
library(pheatmap)
library(randomForest)
library(caret)

set.seed(929)
```

1. Import NOT data
```{r}
expr_df <- read.csv("/Users/kexin/Desktop/project/data/SCs_32_matrix.csv", check.names = FALSE)

metadata <- read.csv("/Users/kexin/Desktop/project/data/SCs_32.csv", check.names = FALSE)

dim(expr_df)
dim(metadata)

head(expr_df[, 1:6])
head(metadata[, 1:6])

cell_labels <- expr_df$cell_label

stopifnot(nrow(expr_df) == nrow(metadata))
```

2. Prepare NOT expression matrix
``` {r, fig.width=6, fig.height=4.5}
expr_mat <- expr_df %>% select(-cell_label)

# Remove Blank probes
expr_mat <- expr_mat %>% select(!starts_with("Blank"))

# Remove genes with missing values
expr_mat <- expr_mat[, colSums(is.na(expr_mat)) == 0]

# Remove genes with zero variance
gene_var <- apply(expr_mat, 2, var)
expr_mat_clean <- expr_mat[, gene_var > 0]

dim(expr_mat_clean)
```

3. Basic QC
```{r}
qc_df <- data.frame(
  cell_label = cell_labels,
  detected_genes = rowSums(expr_mat_clean > 0),
  total_expression = rowSums(expr_mat_clean)
)

ggplot(qc_df, aes(x = detected_genes)) +
  geom_histogram(bins = 30) +
  theme_classic(base_size = 14) +
  labs(
    x = "Number of detected genes",
    y = "Cell count"
  )

ggplot(qc_df, aes(x = total_expression)) +
  geom_histogram(bins = 30) +
  theme_classic(base_size = 14) +
  labs(
    x = "Total expression",
    y = "Cell count"
  )
```

4. PCA
```{r, fig.width=6, fig.height=4.5}
pca_res <- prcomp(expr_mat_clean, center = TRUE, scale. = FALSE)

pca_df <- as.data.frame(pca_res$x)
pca_df$cell_label <- cell_labels

var_explained <- summary(pca_res)$importance[2, ] * 100
cum_var <- cumsum(var_explained)

pc1_lab <- paste0("PC1 (", round(var_explained[1], 1), "%)")
pc2_lab <- paste0("PC2 (", round(var_explained[2], 1), "%)")

ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = pc1_lab,
    y = pc2_lab
  )

scree_df <- data.frame(
  PC_number = seq_along(var_explained),
  Variance = var_explained,
  Cumulative = cum_var
)

ggplot(scree_df[1:30, ], aes(x = PC_number, y = Variance)) +
  geom_point(size = 2) +
  geom_line() +
  theme_classic(base_size = 14) +
  labs(
    x = "Principal component",
    y = "Variance explained (%)"
  )

ggplot(scree_df[1:30, ], aes(x = PC_number, y = Cumulative)) +
  geom_point(size = 2) +
  geom_line() +
  theme_classic(base_size = 14) +
  labs(
    x = "Principal component",
    y = "Cumulative variance explained (%)"
  )
```

5. Add SCs metadata
```{r}
pca_df$sample_id <- metadata$`Sample Id`
pca_df$neurotransmitter <- metadata$Neurotransmitter.Type
pca_df$cell_class <- metadata$Class
pca_df$subclass <- metadata$Subclass
pca_df$supertype <- metadata$Supertype
pca_df$allen_cluster <- metadata$Cluster
pca_df$anatomical_structure <- metadata$Anatomical.Structure
pca_df$anatomical_substructure <- metadata$Anatomical.Substructure

table(pca_df$neurotransmitter)
table(pca_df$subclass)
```

6. PCA colored by biological labels
```{r, fig.width=6, fig.height=4.5}
ggplot(pca_df, aes(x = PC1, y = PC2, color = neurotransmitter)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    color = "Neurotransmitter"
  )

ggplot(pca_df, aes(x = PC1, y = PC2, color = subclass)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    color = "Subclass"
  )

ggplot(pca_df, aes(x = PC1, y = PC2, color = supertype)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    color = "Supertype"
  )
```

7. PCA loading analysis
```{r, fig.width=6, fig.height=4.5}
loadings <- as.data.frame(pca_res$rotation)

pc1_plot <- loadings %>%
  rownames_to_column("gene") %>%
  mutate(abs_loading = abs(PC1)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:20)

ggplot(pc1_plot, aes(x = reorder(gene, abs_loading), y = PC1)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "Gene",
    y = "Loading"
  )

pc2_plot <- loadings %>%
  rownames_to_column("gene") %>%
  mutate(abs_loading = abs(PC2)) %>%
  arrange(desc(abs_loading)) %>%
  slice(1:20)

ggplot(pc2_plot, aes(x = reorder(gene, abs_loading), y = PC2)) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "Gene",
    y = "Loading"
  )
```

8. UMAP without clustering
```{r, fig.width=6, fig.height=4.5}

pc_use <- 1:10

pca_input <- pca_res$x[, pc_use]

umap_res <- uwot::umap(
  pca_input,
  n_neighbors = 10,
  min_dist = 0.3,
  metric = "euclidean"
)

umap_df <- as.data.frame(umap_res)
colnames(umap_df) <- c("UMAP1", "UMAP2")

umap_df$cell_label <- cell_labels
umap_df$sample_id <- metadata$`Sample Id`
umap_df$neurotransmitter <- metadata$Neurotransmitter.Type
umap_df$subclass <- metadata$Subclass
umap_df$supertype <- metadata$Supertype
umap_df$allen_cluster <- metadata$Cluster
umap_df$substructure <- metadata$Anatomical.Substructure

ggplot(umap_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    x = "UMAP1",
    y = "UMAP2"
  )

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = neurotransmitter)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "Neurotransmitter"
  )

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = subclass)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "Subclass"
  )

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = substructure)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_classic(base_size = 14) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "Substructure"
  )
```

```{r}
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = neurotransmitter)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    color = "Neurotransmitter"
  )
```


9. Determine the optimal number of NOT clusters
```{r, fig.width=6, fig.height=4.5}
# Evaluate different k values for k-means clustering.
# Clustering is performed on PCA-reduced expression data.
pca_input <- pca_res$x[, pc_use]

k_results <- data.frame()

for (k in 2:8) {
  
  km <- kmeans(
    pca_input,
    centers = k,
    nstart = 50
  )
  
  sil <- silhouette(km$cluster, dist(pca_input))
  
  k_results <- rbind(
    k_results,
    data.frame(
      k = k,
      silhouette = mean(sil[, 3]),
      total_withinss = km$tot.withinss
    )
  )
}

k_results

ggplot(k_results, aes(x = k, y = silhouette)) +
  geom_point(size = 3) +
  geom_line() +
  theme_classic(base_size = 14) +
  labs(
    x = "Number of clusters",
    y = "Mean silhouette score"
  )

ggplot(k_results, aes(x = k, y = total_withinss)) +
  geom_point(size = 3) +
  geom_line() +
  theme_classic(base_size = 14) +
  labs(
    x = "Number of clusters",
    y = "Total within-cluster sum of squares"
  )
```

10. Final k-means clustering of NOT cells
```{r, fig.width=6, fig.height=4.5}
# Select the k value with the highest silhouette score.
# This provides a data-driven choice of cluster number.
best_k <- 3

km_final <- kmeans(
  pca_input,
  centers = best_k,
  nstart = 50
)

pca_df$cluster_kmeans <- as.factor(km_final$cluster)
umap_df$cluster_kmeans <- pca_df$cluster_kmeans

ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster_kmeans)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = pc1_lab,
    y = pc2_lab,
    color = "Cluster"
  )

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = cluster_kmeans)) +
  geom_point(size = 3, alpha = 0.85) +
  theme_classic(base_size = 14) +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    color = "Cluster"
  )
```

11. Characterization of SCs transcriptomic clusters
```{r}
# Rename k-means cluster labels for easier downstream analysis.
pca_df$cluster <- pca_df$cluster_kmeans
umap_df$cluster <- umap_df$cluster_kmeans

# Check the number of cells in each NOT cluster.
table(pca_df$cluster)
```

12. Neurotransmitter composition of NOT clusters

```{r, fig.width=6, fig.height=4.5}
# Calculate neurotransmitter composition within each transcriptomic cluster.

cluster_neuro <- pca_df %>%
  count(cluster, neurotransmitter) %>%
  group_by(cluster) %>%
  mutate(percent = n / sum(n) * 100)

cluster_neuro

ggplot(
  cluster_neuro,
  aes(x = cluster, y = percent, fill = neurotransmitter)
) +
  geom_col() +
  theme_classic(base_size = 14) +
  labs(
    x = "Transcriptomic cluster",
    y = "Percentage of cells",
    fill = "Neurotransmitter"
  )
```

```{r}
neuro_table <- table(
  pca_df$cluster,
  pca_df$neurotransmitter
)

neuro_table

chisq.test(
  neuro_table,
  simulate.p.value = TRUE,
  B = 10000
)
```

```{r, fig.width=6, fig.height=3}
# Standardized residuals show which subclasses are enriched or depleted
# in each transcriptomic cluster.

chi_subclass_neuro <- chisq.test(
  neuro_table,
  simulate.p.value = TRUE,
  B = 10000
)

stdres_neuro <- chi_subclass_neuro$stdres

round(stdres_neuro, 2)

pheatmap(
  stdres_neuro,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
)
```

14. Association between SCs clusters and Allen subclasses

14.1 Allen subclass composition of NOT clusters
```{r, fig.width=6, fig.height=4.5}
# Keep the top 8 most abundant Allen subclasses and group the rest as "Others".
# This makes the stacked bar plot easier to interpret.

top_subclass <- pca_df %>%
  count(subclass, sort = TRUE) %>%
  slice(1:8) %>%
  pull(subclass)

plot_df <- pca_df %>%
  mutate(
    subclass_plot = ifelse(
      subclass %in% top_subclass,
      subclass,
      "Others"
    )
  ) %>%
  count(cluster, subclass_plot) %>%
  group_by(cluster) %>%
  mutate(percent = n / sum(n) * 100) %>%
  ungroup()

# Order subclasses automatically by overall abundance.
subclass_order <- plot_df %>%
  group_by(subclass_plot) %>%
  summarise(total = sum(n), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(subclass_plot)

# Put "Others" at the end so it appears on top of the stacked bars.
subclass_order <- c(
  setdiff(subclass_order, "Others"),
  "Others"
)

plot_df$subclass_plot <- factor(
  plot_df$subclass_plot,
  levels = subclass_order
)

ggplot(
  plot_df,
  aes(
    x = cluster,
    y = percent,
    fill = subclass_plot
  )
) +
  geom_col() +
  theme_classic(base_size = 14) +
  labs(,
    x = "Transcriptomic cluster",
    y = "Percentage of cells",
    fill = "Allen subclass"
  )
```

```{r}
# Test whether Allen subclass composition differs between NOT clusters.

subclass_table <- table(
  pca_df$cluster,
  pca_df$subclass
)

subclass_table

chisq.test(
  subclass_table,
  simulate.p.value = TRUE,
  B = 10000
)
```

14.2 Standardized residual analysis
```{r, fig.width=6, fig.height=4.5}
# Standardized residuals show which subclasses are enriched or depleted
# in each transcriptomic cluster.

chi_subclass <- chisq.test(
  subclass_table,
  simulate.p.value = TRUE,
  B = 10000
)

stdres <- chi_subclass$stdres

round(stdres, 2)

pheatmap(
  stdres,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  display_numbers = TRUE,
)
```

15. Mean expression differences between NOT clusters
```{r}
# Compare mean gene expression between the two NOT transcriptomic clusters.

expr_cluster <- expr_mat_clean %>%
  as.data.frame() %>%
  mutate(cluster = pca_df$cluster)

cluster_mean <- expr_cluster %>%
  group_by(cluster) %>%
  summarise(across(where(is.numeric), mean))

cluster_mean

cluster_diff <- data.frame(
  gene = colnames(expr_mat_clean),
  mean_cluster1 = as.numeric(cluster_mean[1, -1]),
  mean_cluster2 = as.numeric(cluster_mean[2, -1])
)

# Positive values indicate higher expression in cluster 2.
# Negative values indicate higher expression in cluster 1.
cluster_diff$mean_difference <- cluster_diff$mean_cluster2 - cluster_diff$mean_cluster1

cluster_diff <- cluster_diff %>%
  arrange(desc(abs(mean_difference)))

top_diff_genes <- cluster_diff %>%
  slice(1:20)

top_diff_genes
```

15.1 Genes with the largest mean expression differences between SCs clusters
```{r, fig.width=6, fig.height=4.5}
ggplot(
  top_diff_genes,
  aes(
    x = reorder(gene, abs(mean_difference)),
    y = mean_difference
  )
) +
  geom_col() +
  coord_flip() +
  theme_classic(base_size = 14) +
  labs(
    x = "Gene",
    y = "Mean log2 expression difference\nCluster 2 - Cluster 1"
  )
```

16. Wilcoxon-based differential expression analysis
```{r}
# Identify statistically enriched genes for each cluster
# Each cluster is compared against all other NOT cells

marker_results <- data.frame()

for (cl in levels(pca_df$cluster)) {
  
  cells_in <- pca_df$cluster == cl
  
  for (gene in colnames(expr_mat_clean)) {
    
    expr_in <- expr_mat_clean[cells_in, gene]
    expr_out <- expr_mat_clean[!cells_in, gene]
    
    test <- wilcox.test(expr_in, expr_out)
    
    marker_results <- rbind(
      marker_results,
      data.frame(
        cluster = cl,
        gene = gene,
        mean_in = mean(expr_in),
        mean_out = mean(expr_out),
        mean_difference = mean(expr_in) - mean(expr_out),
        p_value = test$p.value
      )
    )
  }
}

marker_results <- marker_results %>%
  group_by(cluster) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  arrange(cluster, p_adj, desc(abs(mean_difference)))

top_markers <- marker_results %>%
  group_by(cluster) %>%
  arrange(p_adj, desc(abs(mean_difference))) %>%
  slice_head(n = 10)

top_markers
```
```{r}
marker_results <- data.frame()

for (cl in unique(pca_df$anatomical_substructure)) {
  
  cells_in <- pca_df$anatomical_substructure == cl
  
  for (gene in colnames(expr_mat_clean)) {
    
    expr_in <- expr_mat_clean[cells_in, gene]
    expr_out <- expr_mat_clean[!cells_in, gene]
    
    test <- wilcox.test(expr_in, expr_out)
    
    marker_results <- rbind(
      marker_results,
      data.frame(
        anatomical_substructure = cl,
        gene = gene,
        mean_in = mean(expr_in),
        mean_out = mean(expr_out),
        mean_difference = mean(expr_in) - mean(expr_out),
        p_value = test$p.value
      )
    )
  }
}

marker_results <- marker_results %>%
  group_by(anatomical_substructure) %>%
  mutate(p_adj = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  arrange(anatomical_substructure, p_adj, desc(abs(mean_difference)))

top_anatomy_markers <- marker_results %>%
  group_by(anatomical_substructure) %>%
  arrange(p_adj, desc(abs(mean_difference))) %>%
  slice_head(n = 10)

top_anatomy_markers
```

17. Heatmap of SCs cluster marker genes
```{r, fig.width=6, fig.height=4.5}
# Visualize top marker genes across NOT cells.
# Cells are annotated by transcriptomic cluster only to avoid annotation-color mismatch.

top_marker_genes <- unique(top_markers$gene)

heatmap_mat <- expr_mat_clean[, top_marker_genes]

# Scale each gene across cells
heatmap_scaled <- scale(heatmap_mat)

# Add row annotation
annotation_row <- data.frame(
  Cluster = factor(pca_df$cluster)
)

rownames(annotation_row) <- pca_df$cell_label
rownames(heatmap_scaled) <- pca_df$cell_label

# Make sure row order matches
annotation_row <- annotation_row[rownames(heatmap_scaled), , drop = FALSE]


pheatmap(
  heatmap_scaled,
  annotation_row = annotation_row,
  show_rownames = FALSE,
  show_colnames = TRUE,
  fontsize_col = 8,
)
```

18. Known marker gene expression in NOT
```{r, fig.width=6, fig.height=4.5}
# Project selected marker gene expression onto the NOT UMAP.

known_genes <- c("Slc17a6", "Lhfp", "Nrn1", "Nptx2")

known_genes_available <- known_genes[known_genes %in% colnames(expr_mat_clean)]
known_genes_missing <- known_genes[!known_genes %in% colnames(expr_mat_clean)]

known_genes_available
known_genes_missing

for (gene in known_genes_available) {
  
  plot_df <- umap_df
  plot_df$expression <- expr_mat_clean[[gene]]
  
  p <- ggplot(
    plot_df,
    aes(x = UMAP1, y = UMAP2, color = expression)
  ) +
    geom_point(size = 3, alpha = 0.85) +
    theme_classic(base_size = 14) +
    labs(
      x = "UMAP1",
      y = "UMAP2",
      color = "log2 expression"
    )
  
  print(p)
}
```

19. Summary statistics of NOT clusters
```{r}
# Summarize the major transcriptomic and biological features
# of each NOT cluster.

cluster_summary <- pca_df %>%
  group_by(cluster) %>%
  summarise(
    n_cells = n(),
    mean_PC1 = mean(PC1),
    mean_PC2 = mean(PC2),
    mean_PC3 = mean(PC3),
    percent_GABA = mean(neurotransmitter == "GABA") * 100,
    percent_Glut = mean(neurotransmitter == "Glut") * 100,
    percent_GABA_Glyc = mean(neurotransmitter == "GABA-Glyc") * 100
  )

cluster_summary
```

```{r}
cluster_anatomy_table <- table(
  pca_df$cluster,
  pca_df$anatomical_substructure
)

cluster_anatomy_table
```

```{r, fig.width=6, fig.height=4.5}
cluster_anatomy_prop <- as.data.frame(cluster_anatomy_table) %>%
  group_by(Var1) %>%
  mutate(proportion = Freq / sum(Freq))

ggplot(
  cluster_anatomy_prop,
  aes(
    x = Var1,
    y = proportion,
    fill = Var2
  )
) +
  geom_col(position = "fill") +
  theme_classic(base_size = 14) +
  labs(
    x = "Data-driven cluster",
    y = "Proportion",
    fill = "Anatomical substructure"
  )
```

```{r}
library(mclust)

ari <- adjustedRandIndex(
  pca_df$cluster,
  pca_df$anatomical_substructure
)

ari
```

```{r}
library(aricode)

cluster_numeric <- as.integer(
  factor(pca_df$cluster)
)

anatomy_numeric <- as.integer(
  factor(pca_df$anatomical_substructure)
)

nmi_scs <- aricode::NMI(
  cluster_numeric,
  anatomy_numeric
)

nmi_scs
```

```{r}
cluster_to_anatomy <- c(
  "1" = "SCzo",
  "2" = "SCsg",
  "3" = "SCop"
)

pca_df$matched_anatomy <- cluster_to_anatomy[
  as.character(pca_df$cluster)
]

table(
  pca_df$matched_anatomy,
  pca_df$anatomical_substructure
)
```



```{r}
# Match each data-driven cluster to an anatomical substructure

cluster_to_anatomy <- c(
  "1" = "SCzo",
  "2" = "SCsg",
  "3" = "SCop"
)

pca_df$matched_anatomy <- cluster_to_anatomy[
  as.character(pca_df$cluster)
]


# Compare matched clusters with known anatomical substructures

cluster_anatomy_table <- table(
  pca_df$matched_anatomy,
  pca_df$anatomical_substructure
)

cluster_anatomy_table


# Calculate cluster-anatomy concordance

cluster_match_concordance <- mean(
  pca_df$matched_anatomy ==
    pca_df$anatomical_substructure
)

cluster_match_concordance


# Calculate majority-class baseline

anatomy_proportions <- prop.table(
  table(pca_df$anatomical_substructure)
)

anatomy_proportions

baseline_concordance <- max(
  anatomy_proportions
)

baseline_concordance


# Compare cluster-anatomy concordance with the baseline

concordance_difference <-
  cluster_match_concordance - baseline_concordance

cluster_match_concordance
baseline_concordance
concordance_difference
```